\begin{acknowledgement}

I would like to acknowledge the invaluable contributions of those who made this research possible. First and foremost, I express my deepest gratitude to my supervisor, Prof.Amalka Pinidiyaarachchi, for her exceptional guidance and unwavering support throughout this research journey. Without her expertise and mentorship, I could not have achieved this level of work. Whenever I encountered challenges in model development or data analysis, her guidance showed me the path forward. Prof.Pinidiyaarachchi's insightful feedback, technical expertise, and continuous encouragement were the driving force behind this project. Her mentorship not only shaped my research approach and analytical thinking but also instilled in me a deeper appreciation for the intersection of machine learning and agricultural economics. I am truly grateful for her patience, dedication, and commitment to my academic development.

I would like to thank the Department of Statistics and Computer Science, University of Peradeniya, for providing the necessary resources and infrastructure that made this research possible. Access to computational resources was particularly valuable for the intensive training requirements of LSTM neural networks and the extensive data preprocessing pipeline. The department's supportive academic environment fostered the interdisciplinary thinking essential for this work.

I am grateful to the Central Bank of Sri Lanka for providing access to daily market price data, the Copernicus Climate Change Service for precipitation datasets, and the Ceylon Petroleum Corporation for fuel price information. This research would not have been feasible without access to these comprehensive data sources.

I would also like to extend my appreciation to my family and friendst, whose conversations helped me discover my passion for using technology to solve real-world agricultural challenges. Through discussions about the impact of price volatility on farming communities in Sri Lanka, I realized that machine learning could be a greate Artificial Intelligence techniqe for meaningful social impact. Their encouragement motivated me to building an accessible system that farmers, traders, and policymakers could actually use in their daily decision-making, which ultimately shaped the direction of this research.

\end{acknowledgement}