\begin{acknowledgement}

The completion of this research would not have been possible without the support, guidance, and encouragement of numerous individuals and institutions to whom I owe my sincere gratitude.

First and foremost, I express my deepest appreciation to my research supervisor, Dr. [Supervisor Name], Department of Statistics, University of Peradeniya, for the invaluable guidance, constructive criticism, and unwavering support throughout this research journey. Your expertise in statistical modeling and insightful suggestions significantly shaped the direction and quality of this work. The countless hours you dedicated to reviewing drafts, discussing methodological challenges, and providing encouragement during difficult phases were instrumental to this achievement.

I extend my sincere thanks to the academic staff of the Department of Statistics, University of Peradeniya, particularly [Course Coordinator Name] and [Additional Faculty Names], for their continuous support and for providing the foundational knowledge that enabled this research. The rigorous coursework and research methodology training equipped me with the skills necessary to undertake this complex project.

I am grateful to the Central Bank of Sri Lanka for providing access to daily market price data, the Copernicus Climate Change Service for precipitation datasets, the Ceylon Petroleum Corporation for fuel price information, and the Department of Agriculture for agricultural statistics. This research would not have been feasible without access to these comprehensive data sources.

Special thanks to the Dambulla Economic Centre and wholesale market officials who provided contextual insights into market operations, trading patterns, and price determination mechanisms. Your practical knowledge enriched my understanding of the real-world dynamics beyond numerical data.

I acknowledge the technical support received from [IT Support/Computing Resources if applicable], particularly for providing computational infrastructure necessary for training deep learning models and conducting extensive experiments.

My heartfelt gratitude goes to my family—my parents [Names if appropriate] and siblings—for their unconditional love, patience, and encouragement throughout my academic journey. Your sacrifices, understanding during long hours of research, and belief in my capabilities provided the emotional foundation for persevering through challenges.

I am thankful to my friends and colleagues, particularly [Friend/Colleague Names if appropriate], for the stimulating discussions, collaborative learning, moral support, and for making the research process more enjoyable. Your willingness to listen to my ideas, provide feedback, and share in both frustrations and celebrations made this journey less solitary.

Finally, I express gratitude to all farmers, traders, and agricultural stakeholders whose livelihoods motivated this research. It is my hope that this work contributes, even modestly, to improving market efficiency and supporting better-informed decision-making in Sri Lanka's agricultural sector.

While many have contributed to this work's success, I take full responsibility for any errors or shortcomings that remain.

\end{acknowledgement}